\author{
Joseph S. Phillips$^{1,2, \dagger}$ \\
\'{A}rni Einarsson$^{3,4}$ \\ 
Kasha Strickland$^{1}$ \\
Anthony R. Ives$^{2}$ \\
Bjarni K. Kristj\'{a}nsson$^{1}$ \\
Katja R\"{a}s\"{a}nen$^{5,6}$ 
}

\noindent{} 1. Department of Aquaculture and Fish Biology, 
H\'{o}lar University, Sau{\dh}\'{a}rkr\'{o}kur 551 Iceland

\noindent{} 2. Department of Integrative Biology, 
University of Wisconsin, Madison, Wisconsin 53706 USA

\noindent{} 3. M\'{y}vatn Research Station, IS-660 M\'{y}vatn, Iceland

\noindent{} 4. Faculty of Life and Environmental Sciences, 
              University of Iceland, 101 Reykjav\'{i}k, Iceland

\noindent{} 5. Department of Aquatic Ecology, EAWAG and 
Institute of Integrative Biology, ETH Zurich, 
\"{U}berlandstrasse 133, CH-8600 D\"{u}bendorf, Switzerland

\noindent{} 6. Department of Biological and Environmental Sciences, 
University of Jyv\"{a}skyl\"{a}, 40014, Jyv\"{a}skyl\"{a}, Finland

\noindent{} $\dagger$ E-mail: joseph@holar.is

Acknowledgements

Here is a draft of the acknowledgments text (it was not included in the previous drafts because of the double-blind review, but is required for submission). Can you tell me what you think?:

The data used in this manuscript were collected as part of a long-term monitoring program conducted by the M\'{y}vatn Research Station and H\'{o}lar University. We would like to thank Asgrimur Gu{\dh}mundsson, who helped designed the monitoring program. We would also like to thank Antoine Millet, who contributed to data collection for many seasons, along with many other field assistants. This work was supported by RANNIS GOE-195571-052 to BKK and US-NSF LTREB DEB-1556208 to ARI. Author contributions: AE led long-term data collection with contributions from BKK, KR, JSP, and KS. JSP and ARI designed the analysis, and JSP performed the analysis. JSP wrote the first draft of the manuscript, and all authors contributed to subsequent revisions. 
